---
title: "script_1"
format: pdf
editor: visual
---

# Chapter 1

## R as a calculator

R can be used as a calculator. For example, to calculate the square root of the absolute difference between $3^2$ and $4^2$, you can use the following code:

```{r, eval=FALSE}
3^2
exp(2)
log(10)
log10(10)
sqrt(abs(3^2-4^2))
```

## Objects in R

You can assign values to objects in R using `<-` or `=` and calculate with them like with numbers. For example:

```{r, eval=FALSE}
test <- 56
test2 <- 124251762

test + test2
test * test2  
```

Objects in workspace can be listed with `ls()` and removed with `rm()`. For example:

```{r, eval=FALSE}
ls()
rm(test)
```

## Exercise

Solutions to exercises are as follows

```{r, eval=FALSE}
3^2
sqrt(9)
pi^2
sqrt(abs(3^2-4^2))

test <- 3^2
test2 <- 4^2
test3 <- abs(test - test2)
sqrt(test3)

log(exp(4))
log(exp(4), base = exp(1))
log(100, base = 10)

factorial(8)
exp(factorial(3))
```

# Chapter 2

## Functions in R

Functions are called by passing arguments to them. For example:

```{r, eval=FALSE}
mean(5)
mean(x = 5)
mean(y = 5)
mean(x="test")
mean(x = test)
```

Functions can also be defined by the user using the `function` keyword. For example:

```{r, eval=FALSE}
foo <- function(x, y){
  x^2+y^2
}
foo(y=3, x=5)
```

Note that functions should only use the arguments that are defined in the arguments section. Avoid using global variables in functions like this

```{r, eval=FALSE}
foo <- function(x, y){
  x^2+y^2+ test
}
foo(y=3, x=5)
```

## Exercises

1. Formulate the EOQ formula in R
```{r, exe2-1,  echo=T, include=T, collapse=TRUE}
cost_eoq_fun <- function(q, d, co, cl) {
  # returns total cost per period
  # d...demand
  # q...lot size
  # co...ordering cost
  # cl...stock holding cost
  ((1/2)*cl*q)+((d/q)*co)
}

# test cost function
cost_eoq_fun(d=100, q=20, cl = .1, co = 50)

eoq_fun <- function(co, d, cl) {
  # return optimal lot size
  # d...demand
  # co...ordering cost
  # cl...stock holding cost
  sqrt((2*co*d)/cl)
}
# optimal lot size
q.star <-  eoq_fun(d = 100, cl = .1, co = 50)
# optimal cost
cost_eoq_fun(d=100, q=q.star, cl = .1, co = 50)
```

2. Derive a function for calculating weighted Euclidean distance between two points.
```{r, exe2-2,  echo=T, include=T, collapse=TRUE}
weuc_d2_func <- function(x,y,w) {
  # calculates weighted Euclidean distance between x and y
  # y,x...vectors
  # w.. weight vector
  sqrt(sum(w*(x-y)^2))
}
# test distacne function
weuc_d2_func(x = c(1,2,3), y= c(3,2,1), w=c(1,1,1) )
# result should be sqrt(8)
```

Alternatively use intermediate variables within the function

```{r, exe2-2a,  echo=T, include=T, collapse=TRUE}
weuc_d2_func <- function(x,y,w) {
  # calculates weighted Euclidean distance between x and y
  # y,x...vectors
  # w.. weight vector
  tmp.diff <- (x-y)
  tmp.diff.sq <- tmp.diff^2
  tmp.result <- w*sqrt(tmp.diff.sq)
  return(tmp.result)
}
# test distacne function
weuc_d2_func(x = c(1,2,3), y= c(3,2,1), w=c(1,1,1) )
# result should be sqrt(8)
```


3. Formulate a function for the Geometric Poisson density distribution.
```{r, exe2-3,  echo=T, include=T, collapse=TRUE}
geom_pois_dens_fun <- function(n, lambda, theta){
  # calculates density value of geometric Poisson distribution
  # n...integer, demand/successes, theta,lambda..parameters
  k.vec <- 1:n
  sum(exp(-lambda)*lambda^k.vec/factorial(k.vec)*(1-theta)^(n-k.vec)*choose(n-1, k.vec-1))
}
# test function
geom_pois_dens_fun(n=3, lambda=.5, theta = 2)
```



# Chapter 3

Basics on data types and data manipulation

## Create vectors

There are numerous ways to create vectors in R. For example, you can use the `c()` function to concatenate elements into a vector:

```{r, eval=FALSE}
x <- c(1,2,3,4,5)
# not working as expected because different data types
y <- c(1,2,3,5, "test")
y <- 1:5
rep(x = y, times = 10)
x <- rep(x = y,  each = 10)
z <- c(y,x)
zz <- seq(from = 10, to = 35, by =.25)
```

## Create matrices

Matrices can be created by combining vectors or using `matrix()`. For example:

```{r, eval=FALSE}
# matrix with numbers
A <- matrix(1:9, ncol=3)
# matrix with characters
B <- matrix(letters[1:9], ncol=3)
# matrix by column-binding
C <- cbind(c(1,2,3,4,5), c(6,7,8,9,10))
# for data frames similar
df.1 <-  data.frame(zz[1:30], as.character(zz[1:30]))
```

## Indexing vectors and matrices

Vectors and matrices can be indexed using square brackets. For example:

```{r, eval=FALSE}
x <- 1:10
# first element
x[1]
# first five elements
x[1:5]
# last element
x[-1]
# elements greater than 5
x[x>5]

A <- matrix(1:9, ncol=3)
# single element
A[1,2]
# first row
A[1,]
# first column
A[,1]
# first two rows
A[1:2,]
```

## Biathlon data set

The biathlon data set should be imported by "import dataset" menu. Once imported as object `biathlon_results_women` you can access it as follows:

```{r, eval=FALSE}
# View DataFrame
View(biathlon_results_women)

# overview on race types
table(biathlon_results_women$type)

# overview on competition types
table(biathlon_results_women$competition)

# filter for sprint races only
tmp.id.S <- biathlon_results_women$competition == "S"
df.sprint <- biathlon_results_women[tmp.id.S, ]
# mean of total tiem
mean(df.sprint$tot.time)

# filter with tidyverse
library(tidyverse)
biathlon_results_women %>% 
  filter(competition == "S") %>% 
  select(tot.time) %>% 
  unlist() %>% 
  mean()
```
     


## Exercises

1. Calculate the outer product of two vectors (without outer())
```{r, exe3-1,  echo=T, include=T, collapse=TRUE}
x <- 1:5
y <- 10:6
as.matrix(x) %*% t(as.matrix(y))
outer(x,y)
```


2. Define a function that calculates the trace of a matrix.
```{r, exe3-2,  echo=T, include=T, collapse=TRUE}
trace_func <- function(z){
  # calculates trace of z
  # z...matrix
  sum(diag(z))
} 
tmp <- rnorm(9)
A <- matrix(tmp, ncol=3, byrow = T)
trace_func(A)
```


3. Create a vector containing the first 100 Fibonacci numbers.
Most commonly, the Fibonacci numbers are defined recursively by $F_n=F_{n-1}+F_{n-2}$ whereby $F_0 =0$ and $F_1=1$. However, there is also an explicit formulation: $F_n = \sum_{k=0}^{\lfloor \frac{n-1}{2} \rfloor} \binom{n-k-1}{k}$ (check [here](https://en.wikipedia.org/wiki/Fibonacci_number#Mathematics))

```{r, exe3-3,  echo=T, include=T, collapse=TRUE}
fib_num_fun <- function(n){
  # calculate nth Fibonacci number
  # n...number
  k.vec <- 0:floor((n-1)/2)
  sum(choose(n-k.vec-1, k.vec))
}
# vectorize fib_num_fun such that it accepts input vectors
vfib_num_fun <- Vectorize(fib_num_fun)
# doesn't work
fib_num_fun(1:10)
# works
vfib_num_fun(1:100)
```

4. Create a matrix containing the all binominal coefficients up to $n = 50$
```{r, exe3-4,  echo=T, include=T, collapse=TRUE}
pas <- outer(1:10, 1:10, choose)
```

5. Create a list containing
-- vector of 5 small letters
-- vector of 5 capital letters
-- vector of 5 random numbers

```{r, exe3-5-new,  echo=T, include=T, collapse=TRUE}
l1 <- list(small = letters[1:5], capital = LETTERS[1:5], random = rnorm(5))
```

6. Create a matrix with dimension 4×4 and fill it with random numbers.

```{r, exe3-6-new,  echo=T, include=T, collapse=TRUE}
# set random seed
set.seed(123)
A <- matrix(rnorm(16), ncol=4)
```

7. For the matrix generated in task 6, check whether its invertable.

```{r, exe3-7,  echo=T, include=T, collapse=TRUE}
det(A) != 0
```

8. For the matrix generated in task 6, check whether it is symmetric.

```{r, exe3-8,  echo=T, include=T, collapse=TRUE}
A == t(A)
```


